---
title: "Homework 5"
code-annotations: hover
reference-location: margin
execute: 
  message: false
  warning: false
output: 
  html_document:
    highlight-style: a11y-dark
---

### Data
[billboard_top100.csv](../../Lectures/Lecture5/data/billboard_top100.csv)

### Instructions
Answer each of the following questions. Be sure to display **all your code in the rendered** version (use `echo: true` throughout[^longnote]). 

[^longnote]: You can make this a global option for your whole document by putting it directly in the YAML of your qmd:

    ```{YAML}
    ---
    title: "My Document"
    execute:
      echo: true
    ---
    ```
    

```{r}
#| echo: false
knitr::opts_chunk$set(
  comment = ">")
library(tidyverse)
library(ggrepel)
library(ggthemes)
```

### Exercises

1. Download the `billboard` data set introduced in lecture (above) to the same folder where you're saving your qmd for this homework^[If your project directory is different from the directory where you'll be saving these two files, you should run `setwd()` in the **console** and give it a string of the file path to the folder where these two files are saved. This will allow you to run code interactively without error.].
2. Read in the data, clean up the names, and pivot it in a way so the first few rows look like this: 

```{r}
#| echo: false
billboard <- read_csv("billboard_top100.csv") 
billboard_tidy <- billboard |> 
  pivot_longer(starts_with("wk"), 
               names_to = "week", 
               values_to = "rank", 
               values_drop_na = TRUE, 
               names_prefix = "wk", 
               names_transform = list(week = as.integer)) |> 
  janitor::clean_names() 
billboard_tidy
```

3. Create a variable named `date` that corresponds to the `week` based on the `date_entered`^[For instance, if the `date_entered` is `1-13-2000` and `week` is 1, then when `week` is 2 `date` will have a value of `1-20-2000`. Hint: Try using `if_else()` here.]. Read the footnote for a hint but if you need help visualizing what the final dataset might look like, you can reveal another hint below. 

```{r}
#| echo: false
billboard_tidy_date <- billboard_tidy |> 
  mutate(date = if_else(week == 1, 
                       date_entered, 
                       date_entered + weeks(x = week - 1)))
```

::: {.callout-caution collapse="true" icon="false"}
## <span style="color:tomato">{{< fa lightbulb >}}</span> Click for a preview of what the final dataset should look like

```{r}
billboard_tidy_date
```

:::

4. Create a dataset of the song(s) with the <span style="color:#e15759">**most weeks in the top 3 by month of 2000**</span>. The final dataset should look like this:

```{r}
#| echo: false
billboard_top3_month <- billboard_tidy_date |> 
  mutate(month = month(date),
         year = year(date),
         top3 = if_else(rank <= 3 & year == 2000, 1, 0)) |> 
  mutate(peak_weeks = sum(top3), 
         .by = c(month, artist, track)) |> 
  slice_max(peak_weeks, 
            by = month) |> 
  distinct(month, artist, track, peak_weeks) |> 
  arrange(month)
billboard_top3_month
```

5. Pick one month of 2000 and visualize the **entire charting trajectory** of the songs that spent **at least 1 week** in the top 3 during that month. **Hint**: start with the data set created in question 3. An example of what this could look like for April is provided below. 

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 12
#| fig-height: 6
billboard_top3_month_viz <- billboard_tidy_date |> 
  mutate(month = month(date),
         year = year(date),
         top3 = if_else(rank <= 3 & year == 2000, 1, 0)) |> 
  mutate(peak_weeks = sum(top3), 
         month_peak = ifelse(peak_weeks > 0, month, NA),
         .by = c(month, artist, track)) |> 
  filter(any(month_peak == 4), 
         .by = c(track, artist)) 

ggplot(billboard_top3_month_viz, aes(date, rank, group = track, color = artist)) + 
  annotate(geom = "rect", xmin = ymd("2000-04-01"), xmax = ymd("2000-05-01"), ymin = 0, ymax = 85, 
           fill = "#59a14f", alpha = 0.25) + 
  geom_line(show.legend = TRUE) + 
  geom_label_repel(data = billboard_top3_month_viz |> slice_max(date, by = track), 
                   aes(label = track), 
                   show.legend = FALSE) + 
  scale_color_manual("Artist", values = c("#4e79a7","#f28e2c","#e15759","#76b7b2")) + 
  labs(title = "Billboard Top 100 Trajectory for Songs that Hit Top 3 During April 2000", 
       x = "Date", 
       y = "Rank", 
       caption = "Note: April shaded in green") + 
  theme_tufte(base_size = 14) + 
  theme(legend.position = c(0.85, 0.85),
        legend.title.align = 0.5,
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "black", fill = "#f6f7f9"),
        panel.background = element_rect(fill = "#f6f7f9", color = "#f6f7f9"), 
        plot.background = element_rect(fill = "#f6f7f9", color = "#f6f7f9"))
```

> Note: This example plot is an **extremely** specified and polished version of what this can look like. There are a number of ways this can look and you should not be graded nor grade lower for aesthetic features that many of you are still learning. The takeaway here is to challenge yourself to figure out the code to create something that can plot the entire trajectory of the songs that *reached* the top 3 in whatever month you choose. This will require a combination of the skills you've learned in this class thus far. Try this with the content available from lecture first. If you get stuck, there is a hint you can reveal below. 

::: {.callout-caution collapse="true" icon="false"}
## <span style="color:tomato">{{< fa lightbulb >}}</span> Click for skeleton code you can fill in 

Note: This is how **I** did this problem but there are many approaches to every coding puzzle in `R`. If this skeleton code is useful, use it. If not, I'm happy to chat through **your** approach in office hours 🤓 

*Replace all instances of `function`, `variable`, and `value` with what you think the correct answer should be. Additional hints provided by hovering over the code annotation.*

```{r}
#| eval: false
billboard_top3_month_viz <- billboard_tidy_date |> 
  mutate(month = function(variable), # <1>
         year = function(variable), # <2>
         top3 = if_else(variable <= value & variable == value, 1, 0)) |> # <3>
  mutate(month_peak = ifelse(variable > 0, variable, NA), # <4>
         .by = c(month, artist, track)) |> 
  filter(function(month_peak == "value"), # <5>
         .by = c(track, artist)) 

library(ggrepel) # <6>
ggplot(billboard_top3_month_viz, 
       aes(variable, variable, group = variable, color = variable)) + # <7> 
  function + # <8>
  geom_label_repel(data = billboard_top3_month_viz |> function(variable, by = track), # <9>
                   mapping = aes(label = variable)) # <10> 
```

1. What month is associated with each row's chart position?
2. Are there multiple years in this dataset?! Given that we're interested in the Billboard Top 100 for **2000** it might be useful to have a variable that allows us to discriminate between years. 
3. To create some indicator of top 3 status you'll need to provide two conditions (one variable needs to be less than or equal to a certain value and another needs to equal a certain value)
4. Need to create a variable that reflects when a particular song charted in the top 3 and `NA` otherwise
5. In order to get the entire trajectory of a song we can't simply filter for the month when it peaked. Then we'd only be able to plot that snippet of its trajectory. Hint: Which function returns `TRUE` if even 1 element it's given is `TRUE`?
6. Load if you want to add labels
7. You want to visualize the ranking trajectory of a song over time. Hint: `group` is the variable you want to visualize.
8. What geometry would be appropriate here?
9. To properly label this plot you'll need to subset the data, otherwise it will try to plot a label for every date available. 
10. Which variable's text are you trying to label?

:::

::: {.callout-warning icon=false}
## <span style="color:orange">{{< fa exclamation-triangle >}}</span> Before you submit:
Have you remembered to add `embed-resources: true` to your YAML?
:::

### Due Dates

```{r}
#| echo: false
#| message: false
#| warning: false

# reading in base due date schedule
source("../due_dates_schedule.R")

make_due_date_table_fall(5) |> 
  tab_options(table.background.color = "#f6f7f9") |> 
  tab_footnote(
    footnote = md("You will have **two weeks** to complete Homework 5"),
    locations = cells_body(columns = Due_Date, rows = 5)
  ) 
```

