---
title: "Homework 2"
highlight-style: a11y-dark
---

```{r}
#| echo: false
library(gapminder)
```

### Instructions

In this homework, you'll pose a question regarding the Gapminder dataset and investigate it graphically. Rather than using the `gapminder` package like we did in lecture, you'll want to use the `dslabs` package^[Remember to install the package in your console first (not in your qmd file) and then load it with the library() function.], which has a larger subset of the Gapminder data (i.e. more observations and variables).

* At the beginning of your document, write down a research question that is based on the Gapminder dataset (e.g., "How does population change over time in different countries?")
* Create 3-6 plots to answer/investigate your research question. Consider histograms (`geom_histogram()`), scatterplots (`geom_point()`), or lineplots (`geom_line()`).
* Be sure all titles, axes, and legends are clearly labelled (no raw variable names).
* Include at least one plot with `facet_wrap()` or `facet_grid()`.
* You can use other `geoms` like bar charts, box plots, add meaningful vertical or horizontal lines, etc. You may find [**this data visualization cheat sheet**](https://rstudio.github.io/cheatsheets/data-visualization.pdf) helpful.

Your document should be pleasant for a peer to look at, with some organization. You must write up your observations in words as well as showing the graphs. Upload both the `.qmd` file and the `.html` file to Canvas. 

*Optional: If you'd like to compare several specific countries, you can adapt the following pseudo_code below to create a subset of the data with as many countries as you like. Replace `"country1"` et al. with the country name as it appears in the dataset. Use `unique(gapminder$country)` to see a complete list of all the countries available.:*

```{r}
#| eval: false
subset <- gapminder |> 
  filter(country %in% c("county1", "country2", "country3"))
```


### Due Dates

```{r}
#| echo: false
#| message: false
#| warning: false

# reading in base due date schedule
source("../due_dates_schedule.R")

make_due_date_table_spring(2) |> 
  tab_options(table.background.color = "#f6f7f9")
```