---
title: "Overview"
---

This page contains general instructions for assignments/peer reviews. For homework-specific instructions and templates go to the specific assignment listed on the left.

------------------------------------------------------------------------

### Assignment Due Dates

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

library(tidyverse)
library(gt)

schedule <- tibble(Homework = c(1:9), 
                   Due_Date_AA = c("10 October", "17 October", "24 October", "31 October", "7 November", "14 November", "21 November", "28 November", "5 December"), 
                   Peer_Due_Date_AA = c("15 October", "22 October", "29 October", "5 November", "12 November", "19 November", "26 November", "3 November", "10 December"), 
                   Due_Date_AB = c("12 October", "19 October", "26 October", "2 November", "9 November", "16 November", "23 November", "30 November", "7 December"), 
                   Peer_Due_Date_AB = c("17 October", "24 October", "31 October", "7 November", "14 November", "21 November", "28 November", "5 November", "12 December"))

schedule |> 
  gt() |> 
  tab_spanner(label = "Section AA",
              columns = vars(Due_Date_AA, Peer_Due_Date_AA)) |> 
  tab_spanner(label = "Section AB",
              columns = vars(Due_Date_AB, Peer_Due_Date_AB), 
              id = "AB") |> 
  #tab_header(title = md("**Homework & Peer Review Due Dates**")) |> 
  cols_align(align = "center") |> 
  cols_label(Homework = "#",
             Due_Date_AA = "Homework Due",
             Peer_Due_Date_AA = "Peer Review Due",
             Due_Date_AB = "Homework Due",
             Peer_Due_Date_AB = "Peer Review Due") |> 
  tab_style(style = cell_fill(color = "#c7cdac"),
            locations = list(cells_body(columns = c(Due_Date_AA, Due_Date_AB)), 
                             cells_column_labels(columns = c(Due_Date_AA, Due_Date_AB)))) |> 
  tab_style(style = cell_fill(color = "#99a486"),
            locations = list(cells_body(columns = c(Peer_Due_Date_AA, Peer_Due_Date_AB)), 
                             cells_column_labels(columns = c(Peer_Due_Date_AA, Peer_Due_Date_AB)))) |> 
  tab_style(style = cell_fill(color = "#f6f7f9"),
            locations = list(cells_column_spanners(), 
                             cells_body(columns = Homework), 
                             cells_column_labels(columns = Homework))) |> 
  tab_options(table.align = "left", 
              table.font.names = c("Raleway", "Fraunces"), 
              table.border.top.style = "hidden") 
```

------------------------------------------------------------------------

All assignments should be created using Quarto and uploaded to Canvas as both html and qmd files. Your document should be pleasant for a peer to look at and follow all instructions. Use sections for organization, label figures/tables, and use code chunks. 

### **Rubric Students Use to Grade Assignments:**

\

```{r}
#| echo: false
#| warning: false
#| message: false

source("rubric.R")

rubric |> 
  tab_options(table.align = "left", table.font.size = pct(105))
  
```

### **Peer Review:**

You are required to **both grade and provide feedback** on a peer's homework each week. Reading/evaluating the work of others is useful and important! Good programmers realize that most problems have been solved by others; the trick is figuring out how to apply existing code to new problems. You will develop this skill by reading others' code, evaluating their approach to problems, providing suggestions, and pointing out when you learned something new. **Good peer reviews** give constructive, polite, and insightful comments, are specific and concise, use the grading rubric below, and point out what you learned from their code. **Poor peer reviews** are impolite or unprofessional, are generic (i.e., could be applied to anyone's homework), or fail to provide useful suggestions.

::: {.callout-important}
## Late Homework Will Not Be Accepted For Credit
Due to the nature of our schedule, peer reviews are randomly assigned when the due date/time is reached. Therefore, if you don't submit your homework on time, you will not be given a peer's homework to review and vice versa. Life is messy and complicated though and we all miss deadlines for a variety of reasons. Therefore, you can request that I provide you with feedback on a late assignment ({{< var message >}} on Slack) but you won't be able to earn any points for that particular homework. 
:::

### **Rubric Instructor Uses to Grade Peer Review:**

\

```{r}
#| echo: false
#| warning: false
#| message: false

peer_review_rubric |> 
  tab_options(table.align = "left", table.font.size = pct(105))
```


### **When to Contact the Instructor:**

A number of situations may warrant contacting the instructor:

1.  If you believe you have received an unfair peer review, you can **resubmit** your homework for evaluation by the instructor. The assignment will be regraded, which may result in a higher, same, or lower grade.

2.  If you wish for more feedback on a particular part of the homework, you can ask the instructor for additional explanation **without resubmitting** during office hours or over Slack. 

3.  If you receive no feedback, I will provide gentle peer review **on request**.
