title: "CSSS 508 • Lecture 10"
subtitle: "Workflow & Best Practices"
author: "Victoria Sass"
date: "November 14, 2023"
date-format: long
highlight-style:  vim-dark #, oblivion, radical, vim-dark, espresso, zenburn, kate, monochrome, breezedark, haddock
footnotes-hover: true
reference-location: margin
execute: 
  echo: true
  message: false
  warning: false
  fig-format: "svg"
format: 
  revealjs:
    theme: lecture_styles.scss
    slide-number: true
    code-link: true
    chalkboard: true
    incremental: false 
    smaller: true
    preview-links: true
    code-line-numbers: false
    #logo: "../../images/csss-white.png" # need to add css for size <https://stackoverflow.com/questions/74083521/how-to-increase-size-logo-in-quarto-reveal-js-presentation>
---

```{r}
#| echo: false
#| cache: false
require(downlit)
require(xml2)
require(tidyverse)
```

# Roadmap

---

<br>

Last time, we learned about,

*

. . . 

Today, we will cover,

* 

# Code Style

## Naming Variables 

It's good practice to name variables using only lowercase letters, numbers, and `_` (to separate words). 

. . . 

Remember to give them descriptive names, even if that means they're longer. 

. . . 

If you have many related variables, try and be consistent with your naming convention. 
* A common prefix is preferable to a common suffix due to RStudio's autocomplete feature. 

```{r}
#| eval: false
# Code goal: 
short_flights <- flights |> 
  filter(air_time < 60)

# Code foul: 
SHORTFLIGHTS <- flights |> 
  filter(air_time < 60)
```

## Spacing 

For readability you'll want to put spaces around all mathematical operators^[except for ^] (i.e. `+`, `-`, `==`, `<`, etc.) as well as the assignment operator (`<-`). 

```{r}
#| eval: false
# Code goals: 
z <- (a + b)^2 / d
mean(x, na.rm = TRUE)

# Code foul: 
z<-( a + b ) ^ 2/d
mean (x ,na.rm=TRUE)
```

. . . 

To make code easier to skim quickly it's alright to add extra space for better alignment. 

```{r}
#| eval: false
flights |> 
  mutate(
    speed      = distance / air_time,
    dep_hour   = dep_time %/% 100,
    dep_minute = dep_time %%  100
  )
```