---
title: "Homework 3"
highlight-style: a11y-dark
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
```

### Instructions
For this assignment, include all code and relevant output in your rendered html document. 

### Exercises

1. Organize your files for this class in a way that facilitates reproducible research. 
    i) Start with homeworks 1 and 2: make sure the qmds and htmls for each have their own folder within a folder for this class.
    ii) Retroactively turn those homework folders into RStudio Projects by selecting *File* > *New Project...* > *Existing Directory* and selecting the appropriate folder for each homework. Then close each project.
    iii) For the current homework, begin by first creating a project for "Homework3"^[The specific titles used to name your files/folders can be whatever you want as long as they're descriptive and consistent in their naming conventions.].
    iv) Create your .qmd for this homework within this new project^[You'll notice that the upper-right hand corner of your RStudio IDE now displays the name of your new project for this homework. This is your current working directory.]. 
    v) When you're done organizing the folder for this class, write out its current organizational structure (i.e. something similar to the solution to lab question 5[^longnote] with the specific names for files and folders that *you* used). 

[^longnote]: You can embed output-looking text like this example by using a code chunk without the `{r}` like so:

    ```` 
    ```
    CSSS 508
      Homework1/
      ...etc...
    ```
    ````

---

```{r}
x <- rnorm(20, 10, 5)
```

2. Create the vector above:
    i) Look at the help file for `rnorm` and in your own words describe what arguments it takes and what it produces.
    ii) Which arguments are required and which are optional (if any)?
    iii) What does the `r` in `rnorm` stand for? 
    iv) Using `[`, select every value except the last one
    v) Using `[`, select only values that are greater than 10. 

---

```{r}
#| eval: false
ggplot(mpg, aes(x = class)) +
  geom_bar()
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave("mpg-plot.png")
```

3. Run the code above. 
   i) Which of the two plots is saved as mpg-plot.png? Why?^[Hint: Check out the help page!]
   ii) In the files tab of the lower-left pane of your RStudio, look at the project folder for this homework and re-write the current organizational structure for this class with any changes that have occurred (i.e. update your answer to question 1.v.). 

---

4. Go to the [RStudio Tips Twitter account](https://twitter.com/rstudiotips) and find one tip that makes sense to you and looks interesting. Describe it and link to it^[See how to link to websites in markdown syntax [here](https://quarto.org/docs/authoring/markdown-basics.html#links-images)] in your qmd. Maybe practice using it!^[No need to demonstrate this in your qmd.]

---

```{r}
y <- tibble(a = seq(1, 10, 2), 
            b = c("apple", "banana", "strawberry", "peach", "mango"), 
            c = c(rep(TRUE, 3), rep(FALSE, 2)))
```

```{r}
#| echo: false
# seq takes arguments `from`, `to`, and `by` in that order. It creates a vector of numeric values starting with the value input `from`, ending with the value input `by`, in increments of the value input `by`. 

# rep takes a value as its first argument and a integer value, `times` as its default second argument. It repeats the first argument by the specified number of times in the second argument. 
```


5. Create the tibble above. 
   i) What does `seq` do in this example and what arguments does it take? 
   ii) What does `rep` do in this example and what arguments does it take?
   iii) Index `y` using `[` to arrive at the results: 
   
```{r}
#| echo: false
y[c(2, 5), 2]
```
    
<span>&emsp;&emsp;</span> iv) Index `y` using `$` to arrive at the results: 
   
```{r}
#| echo: false
y$c
```
   
<span>&emsp;&emsp;</span> v) What does `y[3:5, ]` produce? What does `:` do?

```{r}
#| echo: false
#| eval: false 
y[3:5, ] # Prints out the 3rd, 4th, and 5th rows
# : is a shortcut for giving a sequence from the first number (on the left) to the second number (on the right) in increments of 1 or -1
```


### Due Dates

```{r}
#| echo: false
#| message: false
#| warning: false

# reading in base due date schedule
source("../../Lectures/due_dates_schedule.R")

make_due_date_table(3) |> 
  tab_options(table.background.color = "#f6f7f9")
```