---
title: "Homework 3"
highlight-style: a11y-dark
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
```

### Instructions
For this assignment, include all code and relevant output. 

### Exercises

1. Organize your files for this class in a way that facilitates reproducible research. 
    i) Start with your homeworks 1 and 2: make sure they each have their own folder within a folder for this class.
    ii) Retroactively turn those homework folders into projects in RStudio by selecting *File* > *New Project...* > *Existing Directory* and selecting the appropriate folder for each homework. Then close each project.
    iii) For the current homework, begin by first creating a project for "Homework3"^[The specific titles used to name your files/folders can be whatever you want as long as they're descriptive and consistent in their naming conventions.].
    iv) Create your .qmd for this homework within this new project^[You'll notice that the upper-right hand corner of your RStudio IDE now displays the name of your new project for this homework. This is your current working directory.]. 
    v) When you're done organizing the folder for this class, write out its current organizational structure (i.e. something similar to the solution to lab question 5[^longnote] with the specific names for files and folders that *you* used). 

[^longnote]: You can embed output-looking text like this example by using a code chunk without the `{r}` like so:

    ```` 
    ```
    CSSS 508
      Homework1/
      
    ```
    ````

---

```{r}
x <- rnorm(20, 10, 5)
```

2. Create the vector above:
    i) Look at the help file for `rnorm` and in your own words describe what arguments it takes and what it produces.
    ii) Which arguments are required and which are optional (if any)?
    iii) What does the `r` in `rnorm` stand for? 
    iv) Using `[`, select every value except the last one
    v) Using `[`, select only values that are greater than 10. 

---

```{r}
#| eval: false
ggplot(mpg, aes(x = class)) +
  geom_bar()
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave("mpg-plot.png")
```

3. Run the code above. 
   i) Which of the two plots is saved as mpg-plot.png? Why?
   ii) Look at the project folder for this homework and re-write the current organizational structure for this class with any changes that have occurred (i.e. update your answer to question 1.v.). 

4. Go to the [RStudio Tips Twitter account](https://twitter.com/rstudiotips) and find one tip that makes sense to you and looks interesting. Describe it, link to it^[See how to link to websites in markdown syntax [here](https://quarto.org/docs/authoring/markdown-basics.html#links-images)] in your qmd, and practice using it!

---

```{r}
y <- tibble(a = seq(1, 10, 2), 
            b = c("apple", "banana", "strawberry", "peach", "mango"), 
            c = c(rep(TRUE, 3), rep(FALSE, 2)))
```

5. Create the tibble above. 
   i) In your own words what does `seq` do and what arguments does it take? 
   ii) In your own words what does `rep` do and what arguments does it take?
   iii) Index `y` using `[` to arrive at the results: 
       ```
       # A tibble: 2 × 1
         b         
         <chr>     
         1 banana    
         2 mango
       ```
    iv) Index `y` using `$` to arrive at the results: 
       ```
       [1]  TRUE  TRUE  TRUE FALSE FALSE
       ```
     v) What does `y[3:5, ]` produce? What do you think `:` does?

### Due Dates

```{r}
#| echo: false
#| message: false
#| warning: false

# reading in base due date schedule
source("../../Lectures/due_dates_schedule.R")

make_due_date_table(3)
```