---
execute: 
  echo: true
  message: false
  warning: false
  fig-format: "svg"
format: 
  revealjs:
    theme: lecture_styles.scss
    highlight-style:  a11y-dark
    reference-location: margin
    slide-number: true
    code-link: true
    chalkboard: true
    incremental: false 
    smaller: true
    preview-links: true
    code-line-numbers: true
    history: false
    progress: true
    link-external-icon: true
    #logo: "../../images/csss-white.png" # need to add css for size <https://stackoverflow.com/questions/74083521/how-to-increase-size-logo-in-quarto-reveal-js-presentation>
---

```{r}
#| echo: false
#| cache: false
require(downlit)
require(xml2)
require(tidyverse)
```

## {#title-slide data-menu-title="Iteration" background="#1e4655" background-image="../../images/csss-logo.png" background-position="center top 5%" background-size="50%"}


[Iteration]{.custom-title}

[CS&SS 508 • Lecture 9]{.custom-subtitle}

[28 November 2023]{.custom-subtitle2}

[Victoria Sass]{.custom-subtitle3}

# Roadmap{.section-title background-color="#99a486"}

---

<br>

Last time, we learned about,

* Importing and exporting data
* Cleaning and reshaping data
* Dates and times

. . . 

Today, we will cover,

* Why Loops?
* `for()` loops
* `while()` loops

# Why Loops?{.section-title background-color="#99a486"}

## Bad Repetition

If someone doesn't know better, they might find the means of variables in the `swiss` data by typing in a line of code for each column:

```{r goofus, eval=FALSE}
mean1 <- mean(swiss$Fertility)
mean2 <- mean(swiss$Agriculture)
mean3 <- mean(swissExamination)
mean4 <- mean(swiss$Fertility)
mean5 <- mean(swiss$Catholic)
mean5 <- mean(swiss$Infant.Mortality)
c(mean1, mean2 mean3, mean4, mean5, man6)
```

Can you spot the problems?

. . . 

How upset would they be if the `swiss` data had 200 columns instead of `r ncol(swiss)`?

## Good Repetition

```{r}
#| include: false
options(digits = 3)
```

You will learn a better way to calculate column means today using loops!

```{r}
means <- rep(NA, ncol(swiss))
for(i in 1:ncol(swiss)){
  means[i] <- mean(swiss[,i])
}
data.frame(Variable=names(swiss),Mean=means)
```

Don't worry about the details yet! 

. . . 

## Don't Repeat Yourself (DRY)!

The **DRY** idea: Computers are much better at doing the same thing over and over again than we are.

::: {incremental}
* Writing code to repeat tasks for us reduces the most common human coding mistakes.
* It also *substantially* reduces the time and effort involved in processing large volumes of data.
* Lastly, compact code is more readable and easier to troubleshoot.
:::

# `for()` Loops{.section-title background-color="#99a486"}

## The `for()` Loop

`for()` loops are the most general kind of *loop*, found in pretty much every programming language.

"**For** each of these values -- in order -- **do this**"

. . . 

Conceptually:

*Given a set of values...*

1. Set an index variable (often `i`) equal to the first value
2. Do something (perhaps depending on `i`)
3. Is there a next value?
   * *YES*: Update to next value, go back to 2.
   * *NO*: Exit loop

We are *looping* through values and repeating some actions.

## `for()` Loop: Diagram

*Given a set of values...*

![](images/loop.svg)

## `for()` Loop: Example

```{r}
for(i in 1:5) {
    # inside for, output won't show up without print()
    print(i^2) 
}
```

Note this runs *5 separate print commands*, which is why each line starts with `[1]`.

## These Do the Same Thing

:::: {.columns}

::: {.column width="50%"}
```{r}
for(i in 1:3) {
    print(i^2) 
}
```
:::

::: {.column width="50%"}
```{r, results='hold'}
i <- 1
print(i^2) 
i <- 2
print(i^2)
i <- 3
print(i^2)
```
:::
::::

## Iteration Conventions

::: {.incremental}
* We call what happens in the loop for a particular value one **iteration**. 
* Iterating over indices `1:n` is *very* common. `n` might be the length of a vector, the number of rows or columns in a matrix or data frame, or the length of a list.
* Common notation: `i` is the object that holds the current value inside the loop.
    + If loops are nested, you will often see `j` and `k` used for the inner loops.
    + This notation is similar to indexing in mathematical symbols (e.g $\sum\limits_{i=1}^n$)
* Note `i` (and `j`,`k`, etc) are just normal objects. You can use any other names you want.
    + Ex: When iterating over rows and/or columns, I often use `row` and/or `col`!
:::

## Iterate Over Characters

What we iterate over doesn't have to be numbers `1:n` or numbers at all! You can also iterate over a character vector in R:

```{r}
for(i in letters[1:3]) {
    print(i)
}
i # in R, this will exist outside of the loop!
```

## Pre-Allocation

Usually in a `for()` loop, you aren't just printing output, but want to store results from calculations in each iteration somewhere.

. . . 

To do that, figure out what you want to store, and **pre-allocate** an object of the right size as a placeholder (typically with missing values as placeholders).

. . . 

Examples of what to pre-allocate based on what you store:

::: {.incremental}
* Single numeric value per iteration: 
  + `rep(NA, num_iter_iters)`
* Numeric vector per iteration: 
  + `matrix(NA, nrow = num_of_iters, ncol = length_of_vector)`
:::

## Pre-Allocation: Numeric

```{r}
iters <- 10 # Set number of interations
output <- rep(NA,iters) # Pre-allocate numeric vector 
output

for(i in 1:iters) { # Run code below iters times
    output[i] <- (i-1)^2 + (i-2)^2
}
output # Display output
```

Steps:

1. Set a number of iterations
2. Pre-allocate a numeric vector of that length
3. Run ten iterations where the output is a mathematical function of each iteration number.

## Pre-Allocation: Numeric Vector per Iteration Matrix

```{r}
rownums <- 3
colnums <- 6
output <- matrix(NA,nrow=rownums,ncol=colnums)

for(i in 1:rownums){
  for(j in 1:colnums){
    output[i,j] <- i + j
  }
}
output
```

## Aside: If/Else Statements

To test a logical statement and then conditionally execute a set of actions, use `if()` and `else`. The structure is:

```{r eval=FALSE}
if(CONDITION){
  SOME CALCULATION
} else{
  A DIFFERENT CALCULATION
}
```

**Warning!** `else` needs to be on same line as the closing brace `}` of previous `if()`.

## If/Else Simple Example


```{r}
if(8  < 10){
  print("Less than 10!")
}else{
  print("Not less than 10!")
}
```

## More Complex If/Else

We can nest together multiple if/elses! if we wish:

```{r}
i <- 13
if(i <= 10) { 
  print("i is less than or equal to 10!")
} else if(i <= 14) {
  print("i is greater than 10, less than or equal to 14")
} else {
  print("i is greater than or equal to 15")
}
```

## Loops with If/Else Statements

Suppose we want to take the numbers between 1 and 5, and divide the evens by 2 and multiply the odds by 2. We could do that using a loop with if/else statements!

```{r}
for(i in 1:5){
  if(i %% 2 == 0){ #check for even numbers
    print(i / 2)
  }else{
    print(i * 2)
  }
}
```

## Handling Special Cases

Aside from the previous toy example, `if()` statements are useful when you have to handle special cases. 

`if()` statements can be used to make a loop ignore or fix problematic cases.

They are also useful for producing error messages, by generating a message *if* an input value is not what is expected.

# `while()` Loops{.section-title background-color="#99a486"}

## `while()`

A lesser-used looping structure is the `while()` loop.

Rather than iterating over a predefined vector, the loop keeps going until some condition is no longer true.

. . . 

Here is the structure:

```{r eval=FALSE}
while(COND IS MET){
  RUN CODE
}
```

If you're not careful, the while loop will run **forever!!**

## Simple `while()` loop example:

```{r}
x <- 0 
while(x < 3){
  x <- x + 1
  print(x)
}
```

What happened in each iteration?

## These Do the Same Thing

:::: {.columns}

::: {.column width="50%"}
```{r}
x <- 0 
while(x < 3){
  x <- x + 1
  print(x)
}
```
:::

::: {.column width="50%"}
```{r, results='hold'}
x <- 0
x <- x+1
print(x)
x <- x+1
print(x)
x <- x+1
print(x)
```
:::
::::

## More Complex Example

Let's see how many times we need to flip a coin to get 4 heads:

```{r}
num_heads <- 0
num_flips <- 0

while(num_heads < 4) {
  # simulating a coin flip
  coin_flip <- rbinom(n = 1, size = 1, prob = 0.5) 

  # keep track of heads
  if (coin_flip == 1) { 
    num_heads <- num_heads + 1 
  }
  
  # update number of coin flips
  num_flips <- num_flips + 1 
}

num_flips # follows negative binomial distribution
```

## Summary

1. Why Loops?
  + To make our lives easier!
2. `for()` loops: 
  + For iterating over a fixed number of items
3. `while()` loops: 
  + For iterating until some condition is met
  
*Let's take a 10 minute break, then return for some activities!*

## Activity 1

1. Create a `for` loop to calculate the maximum value of each variable in the `swiss` data. What are the maximum values of each variable?
2. Using your previous answer as a starting point, create a nested `for` loop to calculate the maximum value for each variable in the `swiss` data (outer loop), and then divide that maximum by 1, 2, and 4 (inner loop). Print the output after each step.
3. Using your previous answer as a starting point, write a loop that does the same calculations as before but stores the values in a matrix with `ncol(swiss)` columns and 3 rows. How will you "pre-allocate" space for the results?

## Activity 1: Answers

Question 1: 

```{r}
for(i in 1:ncol(swiss)) {
  curr_max <- max(swiss[,i])
  print(curr_max)
}
```

## Activity 1: Answers

Question 2: 

```{r}
for(i in 1:ncol(swiss)) {
  curr_max <- max(swiss[,i])
  for(j in c(1,2,4)){
    print(curr_max/j)
  }
}
```

## Activity 1: Answers

Question 3: 

```{r}
results <- matrix(NA, ncol=ncol(swiss),nrow=3)
for(i in 1:ncol(swiss)) {
  curr_max <- max(swiss[,i])
  for(j in 1:3){
    curr_divisor <- c(1,2,4)[j]
    results[j,i] <- curr_max/curr_divisor
  }
}
results
```

## Activity 2


1\. Consider the vector `vec <- c(1,2,NA,3,NA)`. Write a `for` loop that includes an `if`/`else` function so that for each value `x` in `vec`, we print "Missing!" if `x` is NA, and `x^3` otherwise.

2\. What will happen if I run the following loop:

```{r}
#| eval: false
x <- 1
while(x < 10){
  print(x + 1)
}
```

3\. Write a `while()` loop that starts with `x <- 1` and doubles x each iteration, while `x < 100`. Print `x` after each iteration.

## Activity 2: Answers

1\. 

```{r}
for(x in c(1,2,NA,3,NA)){
  if(is.na(x)){
    print("Missing!")
  } else{
    print(x^3)
  }
}
```

## Activity 2: Answers

2\. What will happen if I run the following loop:

```{r}
#| eval: false
x <- 1
while(x < 10){
  print(x + 1)
}
```

  + **Answer:** The while loop will run forever printing `1`, since we are not updating `x`!!

## Activity 2: Answers

3\. Write a `while()` loop that starts with `x <- 1` and doubles x each iteration, while `x < 100`. Print `x` after each iteration.

```{r}
x <- 1
while(x <100){
  x <- x * 2
  print(x)
}
```

*Why does `x` reach 128?!*

# Homework{.section-title background-color="#99a486"}

## Homework 6

HW 6 will be posted on the website shortly! Remember that it is a continuation of HW 5!

## Due dates

```{r}
#| echo: false
#| message: false
#| warning: false

# reading in base due date schedule
source("../due_dates_schedule.R")

make_due_date_table(9)
```