---
title: "Homework 1"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output-ext: html
---

```{r setup}
#| include: false
## set code chunk defaults. Right now, echo = TRUE by default!
knitr::opts_chunk$set(echo = TRUE) 

## Code to load data here

```

This basic R markdown file has some text, inline code, and code chunks! 

# Example Section

## Code Chunks

```{r}
set.seed(1)
norm <- rnorm(100)
hist(x = norm, 
     xlab = "X Axis Name", 
     ylab = "Count", 
     main = "Histogram of 100 observations from a Standard Normal")
```

Unless you include the chunk option `#| echo: false`, your code chunk will be displayed in your final document. 

## Text and Inline Text 

The sample mean of the data plotted above (rounded to 3 decimal points) is `r round(mean(norm), 3)` and the variance is `r round(var(norm), 3)`. 

What's the relationship between standard deviation and variance?

$$ \text{Standard Deviation}(X) = \sqrt{\text{Variance}(X)} $$

The standard deviation of the data plotted above is `r round(sd(norm), 3)`.
