---
highlight-style:  a11y-dark #, oblivion, radical, vim-dark, espresso, zenburn, kate, monochrome, breezedark, haddock
footnotes-hover: true
reference-location: margin
execute: 
  echo: true
  message: false
  warning: false
  fig-format: "svg"
format: 
  revealjs:
    theme: lecture_styles.scss
    slide-number: true
    code-link: true
    chalkboard: true
    incremental: false 
    smaller: true
    preview-links: true
    code-line-numbers: true
    history: false
    progress: true
    #logo: "../../images/csss-white.png" # need to add css for size <https://stackoverflow.com/questions/74083521/how-to-increase-size-logo-in-quarto-reveal-js-presentation>
---

```{r}
#| echo: false
#| cache: false
require(downlit)
require(xml2)
require(tidyverse)
```

## {#title-slide data-menu-title="Writing Functions" background="#1e4655" background-image="../../images/csss-logo.png" background-position="center top 5%" background-size="50%"}

[Writing Functions]{.custom-title}

[CS&SS 508 • Lecture 8]{.custom-subtitle}

[21 November 2023]{.custom-subtitle2}

[Victoria Sass]{.custom-subtitle3}

# Roadmap{.section-title background-color="#99a486"}

---

<br>

Last time, we learned about,

   * Why we use loops
   * `for()` loops
   * `while()` loops

. . . 

Today, we will cover,

   * Aside: Visualizing the Goal
   * Building blocks of functions
   * Simple functions
   * Using functions with `apply()`

# Visualizing the Goal{.section-title background-color="#99a486"}

---

<br>

Before you can write effective code, you need to know *exactly* what you want:

::: {.incremental}
* **Goal:** Do I want a single value? vector? one observation per person? per year?
* **Current State:** What do I currently have? matrix, vector? long or wide format?
* **Translate:** How can I take what I have and turn it into my goal?
  + Sketch out the steps!
  + Break it down into little operations
:::

. . . 

**As we become more advanced coders, this concept is key!!**

**Remember:** *When you're stuck, try searching your problem on Google!!*

# Building blocks of functions{.section-title background-color="#99a486"}

## Why Functions?

R (as well as mathematics in general) is full of functions! 

. . . 

We use functions to:

  * Compute summary statistics (`mean()`, `sd()`, `min()`)
  * Fit models to data (`lm(Fertility~Agriculture,data=swiss)`)
  * Load data (`read_csv()`)
  * Create ggplots (`ggplot()`)
  * And so much more!!
  
## Examples of Existing Functions

::: {.incremental}
* `mean()`:
    + Input: a vector
    + Output: a single number
* `dplyr::filter()`:
    + Input: a data frame, logical conditions
    + Output: a data frame with rows removed using those conditions
* `readr::read_csv()`:
    + Input: a file path, optionally variable names or types
    + Output: a data frame containing info read in from file
:::
  
. . . 

Each function requires **inputs**, and returns **outputs**

## Why Write Your Own Functions?

Functions encapsulate actions you might perform often, such as:

* Given a vector, compute some special summary stats
* Given a vector and definition of "invalid" values, replace with `NA`
* Defining a new logical operator

. . . 

Advanced function applications (not covered in this class):

* Parallel processing
* Generating *other* functions
* Making custom packages containing your functions

## Anatomy of a Function

```{r}
#| eval: false
NAME <- function(ARGUMENT1, ARGUMENT2=DEFAULT){
  BODY
  return(OUTPUT)
}
```

::: {.incremental}
* **Name**: What you call the function so you can use it later
* **Arguments** (aka inputs, parameters): things the user passes to the function that affect how it works
    + e.g. `ARGUMENT1`, `ARGUMENT2`
    + `ARGUMENT2=DEFAULT` is example of setting a default value
    + In this example, `ARGUMENT1`, `ARGUMENT2` values won't exist outside of the function
* **Body**: The actual operations inside the function.
* **Output**: The object inside `return()`. Could be anything (or nothing!)
    + If unspecified, will be the last thing calculated
:::

# Simple functions{.section-title background-color="#99a486"}

## Example 1: Doubling A Number

```{r}
double_x <- function(x){
  double_x <- x * 2
  return(double_x)
}
```

. . . 

Let's run it!

```{r}
double_x(5)
double_x(NA)
double_x(1:2)
```

## Example 2: Extract First/Last

```{r}
first_and_last <- function(x) {
    first <- x[1]
    last  <- x[length(x)]
    return(c("first" = first, "last" = last))
}
```

. . . 

Test it out:

```{r}
first_and_last(c(4, 3, 1, 8))
```

## Example 2: Testing `first_and_last`

What if I give `first_and_last()` a vector of length 1?

```{r}
first_and_last(7)
```

. . . 

Of length 0?

```{r}
first_and_last(numeric(0))
```

. . . 

Maybe we want it to be a little smarter.

## Example 3: Checking Inputs

Let's make sure we get an error message when the vector is too small:

```{r}
#| code-line-numbers: "|3"
smarter_first_and_last <- function(x) {
    if(length(x) < 2){
      stop("Input is not long enough!")
    } else{
      first <- x[1]
      last  <- x[length(x)]
      return(c("first" = first, "last" = last))    
    }
}
```

::: aside
`stop()` ceases running the function and prints the text inside as an error message.
:::

## Example 3: Testing Smarter Function

```{r, error=TRUE}
smarter_first_and_last(NA)
smarter_first_and_last(c(4, 3, 1, 8))
```

. . . 

## Cracking Open Functions

If you type a function name without any parentheses or arguments, you can see its contents:

```{r}
smarter_first_and_last
```

# Using functions with `apply()` {.section-title background-color="#99a486"}

## Applying Functions Multiple Times?

Last week, we saw an example where we wanted to take the mean of each column in the `swiss` data:

```{r}
for(col_index in 1:ncol(swiss)){
  mean_swiss_col <- mean(swiss[,col_index])
  names_swiss_col <- names(swiss)[col_index]
  print(c(names_swiss_col,round(mean_swiss_col,3)))
}
```

*Isn't this kind of complex?!*

## `apply()`, don't loop!

Writing loops can be challenging and prone to bugs!!

. . . 

The `apply()` can solve this issue:

+ **apply** a function to values in each row or column of a matrix
+ Doesn't require preallocation
+ Can take built-in functions or user-created functions.

## Structure of `apply()`

`apply()` takes 3 arguments:

1. Data (a matrix or data frame)
2. Margin (1 applies function to each *row*, 2 applies to each *column*)
3. Function

```{r eval=FALSE}
apply(DATA, MARGIN, FUNCTION)
```

. . . 

For example,

```{r}
apply(swiss, 2, mean)
```

## Example 1

```{r}
row_max <- apply(swiss,1,max) # maximum in each row
head(row_max, 20)
```

## Example 2

```{r}
apply(swiss, 2, summary) # summary of each column
```

**Note:* Matrix output!

## Example 3: User-Created Function

```{r}
scores <- matrix(1:21, nrow=3)
print(scores)
my_function <- function(x){ mean(x+10, na.rm = T) }
apply(scores, 1, my_function)
```

## Activity: Writing A Function

In Olympic diving, a panel of 7 judges provide scores. After removing the worst and best scores, the mean of the remaining scores is given to the diver. We'll write code to calculate this score!

1. Suppose I get you a vector, `x`, of length 7. Write code that will sort the vector from least to greatest, then keep the 2nd-6th elements^[HINT: Use the `sort()` function and square brackets `[ ]` for subsetting].

2. Write a function to calculate a diver's score:
    * Input: Vector of length 7
    * Checks: Check that the vector has length 7 (if not, stop!)
    * Output: Mean score after removing the lowest and greatest scores.
  
3. Calculate the diver's score given `x <- c(2, 1:5, 3)`

## Activity: Solution

1. Sort and extract elements 2 through 6:
    * **Answer:** Given vector `x`, use `sort(x)[2:6]`

2. Function

```{r}
divers_score <- function(x){
  if(length(x) != 7){
    stop("x is not of length 7!")
  } else{
    x_nofirst_nolast <- sort(x)[2:6]
    return(mean(x_nofirst_nolast))
  }
}
```

. . . 

3. Calculate the diver's score given `x <- c(2, 1:5, 3)`

```{r}
divers_score(x = c(2,1:5,3) )
```

## Activity

*These are homework questions!!*

1. Preallocate a matrix of NAs with 3 rows and 8 columns, called `double_matrix`. Manually specify the first column equal to the values 1, 2, and 3. Using a nested loop, fill in the matrix, row by row, such that each value is double that to its left.

2. Write an `apply()` function to take the median value of each column in the `cars` dataset

2. Using `ggplot`, make a scatterplot of the `speed` and `dist` variables in `cars`. Then, add an appropriate horizontal and vertical line symbolizing the median value of each variable^[*Hint:* Using the layers `geom_vline(xintercept = )` and `geom_hline(yintercept = )`].

## Answers

1. Preallocate a matrix of NAs with 3 rows and 8 columns, called `double_matrix`. Manually specify the first column equal to the values 1, 2, and 3. Using a nested loop, fill in the matrix, row by row, such that each value is double that to its left.

```{r}
double_matrix <- matrix(NA, nrow = 3,ncol = 8)
double_matrix[ , 1] <- 1:3
for(row in 1:3){
  for(col in 2:8){
    double_matrix[row, col] <- double_matrix[row, col-1] * 2
  }
}
double_matrix
```

## Answers

2\. Write an `apply()` function to take the median value of each column in the `cars` dataset

```{r}
median_cars <- apply(cars, 2, median)
median_cars
```

## Answers

3\. Make a ggplot

::: {.panel-tabset}

### Code
```{r}
#| eval: false
library(ggplot2)
ggplot(cars, aes(speed, dist)) + 
  geom_point() +
  geom_vline(xintercept = median_cars[1]) +
  geom_hline(yintercept = median_cars[2])
```

### Plot
```{r}
#| echo: false
#| fig-height: 6
#| fig-width: 8
#| fig-align: center
library(ggplot2)
ggplot(cars,aes(speed,dist)) + 
  geom_point() +
  geom_vline(xintercept = median_cars[1]) +
  geom_hline(yintercept = median_cars[2])
```
:::
 
# Homework{.section-title background-color="#99a486"}
 
## Homework 7

Time to work on [Homework 7](https://vsass.github.io/CSSS508/Homework/HW7/homework7.html)!

## Due Dates

* **Section AA:** 
  + Homework #4 due Tuesday, November 21st (before lecture)
  + Peer grading due Sunday, November 26th (by 4:30pm)

<br>

* **Section AB:** 
  + Homework #4 due Thursday, November 23rd (before lab)
  + Peer grading due Tuesday, November 28th (before lecture)
