---
highlight-style:  a11y-dark #, oblivion, radical, vim-dark, espresso, zenburn, kate, monochrome, breezedark, haddock
footnotes-hover: true
reference-location: margin
execute: 
  echo: true
  message: false
  warning: false
  fig-format: "svg"
format: 
  revealjs:
    theme: lecture_styles.scss
    slide-number: true
    code-link: true
    chalkboard: true
    incremental: false 
    smaller: true
    preview-links: true
    code-line-numbers: false
    #logo: "../../images/csss-white.png" # need to add css for size <https://stackoverflow.com/questions/74083521/how-to-increase-size-logo-in-quarto-reveal-js-presentation>
---

```{r}
#| echo: false
#| cache: false
require(downlit)
require(xml2)
require(tidyverse)
options(width = 70)
# knitr::opts_chunk$set(comment = "##")
```

## {#title-slide data-menu-title="Data Structures" background="#1e4655" background-image="../../images/csss-logo.png" background-position="center top 5%" background-size="50%"}


[Data Structures]{.custom-title}

[CS&SS 508 • Lecture 4]{.custom-subtitle}

[24 October 2023]{.custom-subtitle2}

[Victoria Sass]{.custom-subtitle3}

# Preamble: R Data Structures{.section-title background-color="#99a486"}

--- 

<br>

So far, we've been manipulating, summarizing, and making visuals out of data. That's pretty great!!

But now, we need to get more into the weeds of programming...

Today is all about *types of data* in R.

# Roadmap{.section-title background-color="#99a486"}

---

<br>

Last time, we learned about,

* Logical conditions
* Subsetting data
* Modifying data 
* Summarizing data
* Merging together data

. . . 

Today, we will cover,

* Types of Data
* Vectors
* Matrices
* Lists

# Types of Data{.section-title background-color="#99a486"}

+ `numeric`
+ `character`
+ `factor`
+ `logical`
+ `NA`, `NaN`, and `Inf`

## How is my data stored?

Under the hood, R stores different types of data in different ways.

. . . 

+ e.g., R knows that `4.0` is a number, and that `"Vic"` is not a number.

. . . 

So what exactly are the common data types, and how do we know what R is doing?

## Data Types

* **numeric**: `c(1, 10*3, 4, -3.14)`

. . . 

* **character**: `c("red", "blue", "blue")`

. . . 

* **factor**: `factor(c("red", "blue", "blue"))`


. . . 

* **logical**: `c(FALSE, TRUE, TRUE)`

## Note on Factor Vectors

Factors are categorical data that encode a (modest) number of **levels**, like for experimental group or geographic region: 

```{r}
test_group <- factor(c("Treatment", "Placebo", "Placebo", "Treatment"))
test_group
```

. . . 

Why use `factor` instead of `character`? Because `factor` data can go into a statistical model^[Most R models will automatically convert character data to factors. The default reference is chosen alphabetically.].

## Note on Logical Vectors

Remember that `logical` data in R takes on boolean `TRUE` or `FALSE` values.

. . . 

You can do math with logical values, because R makes `TRUE` = 1 and `FALSE` = 0:

```{r}
my_booleans <- c(TRUE, TRUE, FALSE, FALSE, FALSE)
sum(my_booleans)
mean(my_booleans)
```

## Missing or Infinite Data Types

Your data may otherwise be missing or infinite:

1. **Not Applicable** `NA` 

    + Used when data simply is missing or "not available"
    
. . . 

2. **Not a Number** `NaN`

    + Used when you try to perform a bad math operation, e.g., `0 / 0 `
    
. . . 

3. **Infinite** `Inf`, `-Inf`

    + Used when you divide by 0, e.g., `-5/0` or `5/0`

## Checking Data Types

`class()` tells us what type of data we have:

```{r}
class4     <- class(4)
classAB    <- class(c("A", "B"))
classABFac <- class(factor("A", "B"))
classTRUE  <- class(TRUE)

c(class4, classAB, classABFac, classTRUE)
```

## Testing Data Types

There are also functions to **test** for certain data types:

```{r}
c(is.numeric(5),  is.character("A"))

is.logical(TRUE)

c(is.infinite(-Inf),  is.na(NA),  is.nan(NaN))
```

Warning: `NA` is not `NaN`!!!

# Vectors{.section-title background-color="#99a486"}

+ Creating Vectors
+ Vector Math
+ Subsetting Vectors

## Making Vectors

In R, we call a set of values of the same type a **vector**. We can create vectors using the `c()` function ("c" for **c**ombine or **c**oncatenate).

```{r}
c(1, 3, 7, -0.5)
```

. . . 

Vectors have one dimension: **length**

```{r}
length(c(1, 3, 7, -0.5))
```

. . . 

All elements of a vector are the same type (e.g. numeric or character)!

If you mix character and numeric data, it will convert everything to *characters*!

## Generating Numeric Vectors

There are shortcuts for generating numeric vectors:

```{r}
1:10
```

. . . 

```{r}
seq(-3, 6, by = 1.75) # Sequence from -3 to 6, increments of 1.75
```

. . . 

```{r}
rep(c(0, 1), times = 3) # Repeat c(0,1) 3 times
rep(c(0, 1), each = 3) # Repeat each element 3 times
```

## Element-wise Vector Math

When doing arithmetic operations on vectors, R handles these *element-wise*:

```{r}
c(1, 2, 3) + c(4, 5, 6)
c(1, 2, 3, 4)^3 # exponentiation with ^
```

Common operations: `*`, `/`, `exp()` = $e^x$, `log()` = $\log_e(x)$

## Vector Recycling

If we work with vectors of different lengths, R will **recycle** the shorter one by repeating it to make it match up with the longer one:
```{r}
c(0.5, 3) * c(1, 2, 3, 4)
c(0.5, 3, 0.5, 3) * c(1, 2, 3, 4) # same thing
```

## Scalars as Recycling

A special case of recycling involves arithmetic with **scalars** (a single number). These are vectors of length 1 that are recycled to make a longer vector:
```{r}
3 * c(-1, 0, 1, 2) + 1
```

## Warning on Recycling

Recycling doesn't work so well with vectors of incommensurate lengths:

```{r}
#| warning: true
c(1,2) + c(100,200,300)
```

Be careful!!

## Example: Standardizing Data

Let's say we had some test scores and we wanted to put these on a standardized scale: 

$$z_i = \frac{x_i - \text{mean}(x)}{\text{SD}(x)}$$

. . . 

```{r}
x <- c(97, 68, 75, 77, 69, 81)
z <- (x - mean(x)) / sd(x)
round(z, 2)
```


## Math with Missing Values

Even one `NA` "poisons the well": You'll get `NA` out of your calculations unless you add the extra argument `na.rm = TRUE` (availabe in some functions):

. . . 

```{r}
vector_w_missing <- c(1, 2, NA, 4, 5, 6, NA)
mean(vector_w_missing)
mean(vector_w_missing, na.rm = TRUE)
```

## Subsetting Vectors

We can **subset** a vector in a number of ways:

* Passing a single index or vector of entries to **keep**:

```{r}
first_names <- c("Andre", "Brady", "Cecilia", "Danni", "Edgar", "Francie")
first_names[1]
first_names[c(1, 2)]
```

. . . 

* Passing a single index or vector of entries to **drop**:

```{r}
first_names[-3]
```

# Matrices{.section-title background-color="#99a486"}

## Matrices: Two Dimensions

**Matrices** extend vectors to two **dimensions**: **rows** and **columns**. We can construct them directly using `matrix`.

R fills in a matrix column-by-column (**not** row-by-row!)

```{r}
a_matrix <- matrix(first_names, nrow = 2, ncol = 3)
a_matrix
```

## Binding Vectors

We can also make matrices by *binding* vectors together with `rbind()` (**r**ow **bind**) and `cbind()` (**c**olumn **bind**).

```{r}
b_matrix <- cbind(c(1, 2), c(3, 4), c(5, 6))
b_matrix
c_matrix <- rbind(c(1, 2, 3), c(4, 5, 6))
c_matrix
```

## Subsetting Matrices

We subset matrices using the same methods as with vectors, except we index them with `[rows, columns]`:

```{r}
a_matrix[1, 2] # row 1, column 2
a_matrix[1, c(2,3)] # row 1, columns 2 and 3
```

. . . 

We can obtain the dimensions of a matrix using `dim()`.

```{r}
dim(a_matrix)
```

## Matrices Becoming Vectors

If a matrix ends up having just one row or column after subsetting, by default R will make it into a vector. 

```{r}
a_matrix[, 1] 
```

. . . 

You can prevent this behavior using `drop = FALSE`.

```{r}
a_matrix[, 1, drop = FALSE] 
```

## Matrix Data Type Warning

Matrices can contain numeric, integer, factor, character, or logical. But just like vectors, *all elements must be the same data type*.

```{r}
bad_matrix <- cbind(1:2, c("Victoria", "Sass"))
bad_matrix
```

In this case, everything was converted to characters!

## Matrix Dimension Names

We can access dimension names or name them ourselves:

```{r}
rownames(bad_matrix) <- c("First", "Last")
colnames(bad_matrix) <- c("Number", "Name")
bad_matrix
bad_matrix[ ,"Name", drop = FALSE]
```

## Matrix Arithmetic

Matrices of the same dimensions can have math performed entry-wise with the usual arithmetic operators:

```{r}
matrix(c(2, 4, 6, 8),nrow = 2, ncol = 2) / matrix(c(2, 1, 3, 1),nrow = 2, ncol = 2)
```

## "Proper" Matrix Math

To do matrix transpositions, use `t()`.

```{r}
e_matrix <- t(c_matrix)
e_matrix
```

. . . 

To do actual matrix multiplication (not entry-wise), use `%*%`.

```{r}
f_matrix <- c_matrix %*% e_matrix
f_matrix
```

## "Proper" Matrix Math (cont.)

To invert an invertible square matrix, use `solve()`.

```{r}
g_matrix <- solve(f_matrix)
g_matrix
```

## Matrices vs. data.frames and tibbles

All of these structures display data in two dimensions

+ `matrix`

    + Base R
    + Single data type allowed

. . . 

+ `data.frame`

    + Base R (default for data storage)
    + Stores multiple data types
    
. . . 

+ `tibbles`

    + tidyverse
    + Stores multiple data types
    + Displays nicely

. . . 

In practice, data.frames and tibbles are very similar!

## Creating `data.frame`s

We create a `data.frame` by specifying the columns separately:

```{r}
data.frame(Column1Name = c(1,2,3),
           Column2Name = c("A","B","C"))
```

*Note:* `data.frame`s allow for **mixed data types!**

# Lists{.section-title background-color="#99a486"}

## What are Lists?

**Lists** are objects that can store multiple types of data.


```{r}
my_list <- list("first_thing"  = 1:5,
                 "second_thing" = matrix(8:11, nrow = 2))
my_list
```

## Accessing List Elements

You can access a list element by its name or number in `[[ ]]`, or a `$` followed by its name:

```{r}
my_list[["first_thing"]]
my_list[[1]]
my_list$first_thing
```

## Why Two Brackets `[[` `]]`?

Double brackets get *the actual element* — as whatever data type it is stored as, in that location in the list.

```{r}
str(my_list[[1]])
```

. . . 

If you use single brackets to access list elements, you get a **list** back. 

```{r}
str(my_list[1])
```

## `names()` and List Elements

You can use `names()` to get a vector of list element names:

```{r}
names(my_list)
```

## Example: Regression Output

When you perform linear regression in R, the output is a list!

```{r}
lm_output <- lm(speed ~ dist, data = cars)
is.list(lm_output)
names(lm_output)
lm_output$coefficients
```

# Summary{.section-title background-color="#99a486"}

1. Types of Data: `numeric`, `character`, `factor`, `logical`, `NA`, `NaN`, `Inf`
1. Vectors: `c()`
1. Matrices: `matrix`, `data.frame`, `tibble`
1. Lists: `list`

*Let's take a 10 minute break, then come back together for some practice!*

## Mini-Check 1: Types of Data

In each case, what will R return?

::: {.incremental}
* `is.numeric(3.14)`
  + `TRUE`
* `is.numeric(pi)`
  + `TRUE`
* `is.logical(FALSE)`
  + `TRUE`
* `is.nan(NA)`
  + `FALSE`
:::
    
## Mini-Check 2: Vectors

::: {.incremental}
* What does `sum(c(1, 2, NA))` output?
  + `NA`. The code `sum(c(1, 2, NA), na.rm = TRUE)` would output `3`.
:::

<br>

::: {.incremental}
* What does `rep(c(0, 1), times = 2)` output?
  + `c(0, 1, 0, 1)`
:::

<br>

::: {.incremental}
* I want to get the first and second elements of my vector, `a_vector`. What's wrong with the code `a_vector[1, 2]` ?
  + `a_vector[c(1, 2)]`
:::

## Activity: Matrices and Lists

1. Write code to create the following matrix:

```{r echo=FALSE}
matrix(c("A", "B", "C", "D", "E", "F"), nrow = 2, byrow = TRUE)
```

2. Write a line of code to extract the second column. Ensure the output is still a *matrix*.

```{r echo=FALSE}
matrix(c("A", "B", "C", "D", "E", "F"), nrow = 2, byrow = TRUE)[ ,2, drop = FALSE]
```

3. Complete the following sentence: "Lists are to vectors, what data frames are to..."

4. Create a list that contains 3 elements: 
    i) "ten_numbers" (integers between 1 and 10)
    ii) "my_name" (your name as a character) 
    iii) "booleans" (vector of `TRUE` and `FALSE` alternating three times)

## Answers

1\. Write code to create the following matrix:

```{r}
mat_test <- matrix(c("A", "B", "C", "D", "E", "F"), nrow = 2, byrow = TRUE)
mat_test
```

. . . 

2\. Write a line of code to extract the second column. Ensure the output is still a *matrix*.

```{r}
mat_test[ ,2, drop = FALSE]
```

## Answers

3\. Complete the following sentence: "Lists are to vectors, what data frames are to...**Matrices!**^[*Lists and data frames can contain mixed data types, while vectors and matrices can only contain one data type.*]" 

. . . 

4\. Create a list that contains 3 elements:

```{r}
my_new_list <- list("ten_numbers" = 1:10,
                    "my_name" = "Victoria Sass",
                    "booleans" = rep(c(TRUE, FALSE), times = 3))
my_new_list
```

# Homework{.section-title background-color="#99a486"}

## Homework 4

For Homework 4, you will fill in an Quarto template on our [course website](https://vsass.github.io/CSSS508/Homework/HW4/homework4.html) that walks you through the process of creating, accessing, and manipulating R data structures. Enter values in the qmd document and Render it to check your answers!

+ *Render after entering each answer!!*  If you get an error, check to see if undoing your last edit solves the problem. Coding an assignment to handle all possible mistakes is really hard! 

+ This assignment is *long*, so *start early*.

On the due date, I will provide a key for the written answers. You will grade those answers as part of your peer review. In addition, you'll be asked to comment on the style of your peer's code and what you yourself did similarly/different. Please remember to provide a numerical grade (0-3), as always.

## Due Dates

* **Section AA:** 
  + Homework #4 due Tuesday, October 31st (before lecture)
  + Peer grading due Sunday, November 5th (by 4:30pm)

<br>

* **Section AB:** 
  + Homework #4 due Thursday, November 2nd (before lab)
  + Peer grading due Tuesday, November 7th (before lecture)


