---
execute: 
  echo: true
  message: false
  warning: false
  fig-format: "svg"
format: 
  revealjs:
    highlight-style:  a11y-dark
    reference-location: margin
    theme: lecture_styles.scss
    slide-number: true
    code-link: true
    chalkboard: true
    incremental: false 
    smaller: true
    preview-links: true
    code-line-numbers: true
    history: false
    progress: true
    link-external-icon: true
---

```{r}
#| echo: false
#| cache: false
require(downlit)
require(xml2)
require(tidyverse)
```

## {#title-slide data-menu-title="Working with Text Data" background="#1e4655" background-image="../../images/csss-logo.png" background-position="center top 5%" background-size="50%"}

[Working with Text Data]{.custom-title}

[CS&SS 508 • Lecture 7]{.custom-subtitle}

[14 November 2023]{.custom-subtitle2}

[Victoria Sass]{.custom-subtitle3}

# Roadmap{.section-title background-color="#99a486"}

---

<br>

Last time, we learned about,

* Aside: Visualizing the Goal
* Building blocks of functions
* Simple functions
* Using functions with `apply()`

. . . 

Today, we will cover,

* Basics of Strings
* Strings in Base R
* Strings in `stringr` (tidyverse)

# Basics of Strings{.section-title background-color="#99a486"}

## Basics of Strings

::: {.incremental}
* A general programming term for a unit of character data is a **string**
  + Strings are a *sequence of characters*
  + In R, "strings" and "character data" are mostly interchangeable.
  + Some languages have more precise distinctions, but we won't worry about that here!
:::

::: {.incremental}
* We can create strings by surrounding text, numbers, spaces, or symbols with quotes!
  + Examples: `"Hello! My name is Vic"` or `"%*$#01234"`
:::

## Basics of Strings 

R can treat strings in funny ways!

```{r}
"01" == "1"
"01" == 1
"1" == 1
```

. . . 

*Reminder:* We can check **data types** using the `class()` function!

```{r}
c(class("1"), class(1))
```

# Strings in Base R{.section-title background-color="#99a486"}

* `nchar()`
* `substr()`
* `paste()`

## Data: King County Restaurant Inspections!

Today we'll study real data on **food safety inspections in King County**, collected from [data.kingcounty.gov](https://data.kingcounty.gov/Health/Food-Establishment-Inspection-Data/f29f-zza5).

Note these data are *fairly large*. The following code can be used to download the data directly from my Github page:

```{r}
restaurants <- read_csv(file = "https://raw.githubusercontent.com/vsass/CSSS508/main/Lectures/Lecture8/data/restaurants.csv")
```

## Quick Examination of the Data

```{r}
names(restaurants)
dim(restaurants)
```

## Quick Examination of the Data

**Good Questions to Ask:**

* What does each row represent?
* Is the data in long or wide format?
* What are the key variables?
* How are the data stored? (*data type*)

## `nchar()`

The `nchar()` function calculates the *number of characters* in a given string. 

* `length()` doesn't work with strings!!
  * Why not?

. . . 

```{r}
nchar("Victoria Sass")
length("Victoria Sass")
```

. . . 

In our `restaurants` data, let's see how many characters are in each zip code:

```{r nchar_zip}
length_zip <- nchar(restaurants$`Zip Code`) # notice when a variable name has spaces we need to contain it within backticks so R knows it is a singular entity
table(length_zip)
```

. . . 

## `substr()`

The `substr()` function allows us to extract characters from a string.

. . . 

For example, we can extract the third through fifth elements of a string as follows:

```{r}
substr("98126", 3, 5)
```

## `substr()`

Let's extract the first five characters from each zip code in the restaurants data, and add it to our dataset.

```{r}
library(dplyr)
restaurants$ZIP_5 <- substr(restaurants$`Zip Code`, 1, 5) 
restaurants %>% 
  distinct(ZIP_5) %>% 
  head()
```

## `paste()`

We combine strings together using `paste()`. By default, it puts a space between different strings.

. . . 

For example, we can combine `"Victoria"` and `"Sass"` as follows:

```{r}
paste("Victoria", "Sass")
```

## More complex `paste()` commands

There are two additional common arguments to use with `paste()`:

1. `sep =` controls what separates vectors, entry-wise
1. `collapse =` controls if/how multiple outputs are collapsed into a single string.

. . . 

Examples:

```{r}
paste("CSSS", "508", sep = "_")
paste(c("CSSS", "STAT"), "508", sep = "_")
paste(c("CSSS", "STAT"), "508", sep = "_", collapse = " , ")
```

*When do we get one string as output vs. two?*

## `paste()` Example

Let's use `paste()` to create complete mailing addresses for each restaurant:

. . . 

```{r mailing_address}
restaurants$mailing_address <- paste(restaurants$Address, ", ", restaurants$City, 
                                     ", WA ", restaurants$ZIP_5, sep = "")
restaurants %>% 
  distinct(mailing_address) %>% 
  head()
```

# Strings in `stringr` {.section-title background-color="#99a486"}

* `str_length()`
* `str_sub()`
* `str_c()`
* `str_to_upper()`, `str_to_lower()`, and `str_to_title()`
* `str_trim()`
* `str_detect()`
* `str_replace()`

## `stringr`

`stringr` is yet another R package from the Tidyverse (like `ggplot2`, `dplyr`, `tidyr`, `lubridate`, `readr`).

. . . 

It provides TONS of functions for working with strings:

* Some are equivalent/better versions of Base R functions
* Some can do *fancier* tricks with strings
  
. . . 

*Most* `stringr` functions begin with "`str_`" to make RStudio auto-complete more useful.

. . . 

We'll cover the basics today, but know there's much more out there!
    
```{r load_stringr}
library(stringr)
```

## Equivalencies: `str_length()`

`str_length()` is equivalent to `nchar()`:

```{r nchar_v_str_length}
nchar("weasels")
str_length("weasels")
```

. . . 

## Equivalencies: `str_sub()`

`str_sub()` is like `substr()`:

```{r}
str_sub("Washington", 2, 4)
```

. . . 

`str_sub()` also lets you put in negative values to count backwards from the end (-1 is the end, -3 is third from end):

```{r str_sub_example}
str_sub("Washington", 4, -3)
```

## Equivalencies: `str_c()`

`str_c()` ("string combine") is just like `paste()` but where the default is `sep = ""` (no space!)

```{r str_c_example}
str_c(c("CSSS","STAT"), 508)
str_c(c("CSSS","STAT"), 508, sep = " ")
str_c(c("CSSS","STAT"), 508, sep = " ", collapse = ", ")
```

## Changing Cases

`str_to_upper()`, `str_to_lower()`, `str_to_title()` convert cases, which is often a good idea to do before searching for values:

```{r make_seattle_uppercase}
unique_cities <- unique(restaurants$City)
unique_cities %>% 
  head()

str_to_upper(unique_cities) %>% 
  head()
str_to_lower(unique_cities) %>% 
  head()
str_to_title(unique_cities) %>% 
  head()
```

## Whitespace: `str_trim()`

Extra leading or trailing whitespace is common in text data:

```{r show_whitespace}
unique_names <- unique(restaurants$Name)
unique_names %>% 
  head(3)
```

We can remove the whitespace using `str_trim()`:

```{r clean_whitespace}
str_trim(unique_names) %>% 
  head(3)
```

## Patterns!

It's common to want to see if a string satisfies a certain *pattern*.

. . . 

We did this with numeric values earlier in this course!

```{r}
cars %>% 
  filter(speed < 5 | speed > 24)
cars %>% 
  filter(dist > 2 & dist <= 10)
```

## Patterns: `str_detect()`

We can do similar pattern-checking using `str_detect()`:

```{r}
#| eval: false
str_detect(string, pattern)
```

* `string` is the character string (or vector of strings) we want to examine
* `pattern` is the pattern that we're checking for inside `string`
* Output: TRUE/FALSE vector indicating if pattern was found

. . . 

```{r}
str_detect(string = c("Hello", "my middle name", "is Anne"),
           pattern = "a")
str_detect(string = c("Hello", "my middle name", "is Anne"),
           pattern = "A")
```

Results are case-sensitive!!

## Patterns: `str_detect()`

Let's see which phone numbers are in the 206 area code:

```{r}
unique_phones <- unique(restaurants$Phone)
unique_phones %>% tail(4)
str_detect(unique_phones, "206") %>% tail(4)
```

## Replacement: `str_replace()`

What about if you want to replace a string with something else? Use `str_replace()`!

. . . 

This function works very similarly to `str_detect()`, but with one extra argument:

```{r eval=FALSE}
str_replace(string, pattern, replacement)
```

* `replacement` is what `pattern` is substituted for.

. . . 

```{r}
str_replace(string = "Hi, I'm Vic",
            pattern = "Hi", replacement = "Hello")
```

## Replacement: `str_replace()`

In the `Date` variable, let's replace each dash ("-") with an underscore ("_")

```{r}
dates <- restaurants$`Inspection Date`
dates %>% 
  tail(3)
str_replace(dates, "/", "_") %>% 
  tail(3)
```

Wait, what?

## Replacement: `str_replace_all()`

`str_replace()` only changes the **first** instance of a pattern in each string! 

. . . 

If we want to replace **all** patterns, use `str_replace_all()`

```{r}
dates <- restaurants$`Inspection Date`
dates %>% tail(3)
str_replace_all(dates, "/", "_") %>% 
  tail(3)
```

## Quick Summary

We've seen lots of functions today! 

*Don't try to memorize them!* Instead, use this page as a reference.

* Character Length: `nchar` and `str_length`
* Subsetting: `substr` and `str_sub`
* Combining: `paste` and `str_c`
* Case Changes: `str_to_upper()`, `str_to_lower()`, and `str_to_title()`
* Removing Whitespace: `str_trim`
* Pattern Detection/Replacement: `str_detect()` and `str_replace()`

## Activity 1: Base R Functions

The variable `Inspection Date` is in the format "MM/DD/YYYY". In this question, we'll change the format using functions for strings.

1. How long is each character string in this variable?
2. Use `substr()` to extract the month of each entry and save it to an object called "months"
3. Use `substr()` to extract the year of each entry and save it to an object called "years"
4. Use `paste()` to combine each month and year, separated by an underscore (`_`). Save this as a new variable in the data called "Inspection_Date_Formatted"

## Activity: Answers

The variable `Inspection_Date` is in the format "MM/DD/YYYY". In this question, we'll change the format using functions for strings.

1. How long is each character string in this variable?

```{r}
table(nchar(restaurants$`Inspection Date`))
```

. . . 

2. Use `substr()` to extract the month of each entry and save it to an object called "months"

```{r}
months <- substr(restaurants$`Inspection Date`, 1, 2)
months |> 
  head(5)
```

## Activity: Answers

3. Use `substr()` to extract the year of each entry and save it to an object called "years"

```{r}
years <- substr(restaurants$`Inspection Date`, 7, 10)
years |> 
  head(5)
```

. . . 

4. Use `paste()` to combine each month and year, separated by an underscore (`_`). Save this as a new variable in the data called "Inspection_Date_Formatted"

```{r}
restaurants$Inspection_Date_Formatted <- paste(months, years, sep = "_")

restaurants %>% 
  select(Name, `Inspection Date`, Inspection_Date_Formatted) %>%
  head(5)
```

# Dates and Times {background-image="images/lubridate.png" background-size="contain" background-opacity="0.25" background-position="center" .section-title background-color="#99a486"}

## `lubridate`

The package `lubridate` (part of the tidyverse!) has a *very large* number of functions you can use!

. . . 

* Converting dates/times between formats (DD-MM-YY to YY-MM-DD)
* Extracting dates/times (day of week, month, leap years, etc.)
* Math with dates/times (time zone conversions, etc.)

. . . 

There's too much to cover all of it, but I'll run through a few examples.

## Dates in `billboard_2000` 

```{r}
#| echo: false
billboard_2000_raw <- read_csv(file = "https://raw.githubusercontent.com/vsass/CSSS508/main/Lectures/Lecture5/data/billboard.csv")
billboard_2000 <- billboard_2000_raw %>%
  pivot_longer(cols=wk1:wk76,
               names_to ="week",
               values_to = "rank")
```


```{r}
billboard_2000 %>% 
  select(date.entered) %>% 
  head(10)
```

## Extracting Year, Month, or Day

```{r }
library(lubridate)
head(billboard_2000$date.entered, 5)
year(billboard_2000$date.entered) %>% 
  head(5)
month(billboard_2000$date.entered) %>% 
  head(5)
day(billboard_2000$date.entered) %>% 
  head(5)
```

## Extracting Weekday, Quarter, and Leap Year Boolean

```{r}
wday(billboard_2000$date.entered) %>% 
  head(5)
quarter(billboard_2000$date.entered) %>% 
  head(5)
leap_year(billboard_2000$date.entered) %>% 
   head(5)
```


## Activity 2: HW 8

Let's examine the coffee shops of King County!

1. Filter your data to only include rows in which the `Name` includes the word "coffee" (in any case!)

2.. Create a new variable in your data which includes the length of the business name, after removing beginning/trailing whitespace.

3. Create a new variable in your data for the inspection year, *using a `stringr` function!*

4. Create side-by-side boxplots for the length of business name vs. year.

5. Calculate the maximum `Inspection Score` by business and year.

6. Create a line plot of maximum score ("MaxScore") over time ("Year"), by business ("Name"). That is, you should have a single line for each business. (Don't try to label them, as there are far too many!)

## Activity: Solutions

1\. Filter your data to only include rows in which the `Name` includes the word "coffee" (in any case!)

```{r}
coffee <- restaurants
coffee$Name <- str_to_lower(coffee$Name)
coffee <- coffee %>% 
  filter(str_detect(Name, "coffee"))
coffee |> 
  head(5)
```

## Activity: Solutions

2\.Create a new variable in your data which includes the length of the business name, after removing beginning/trailing whitespace.

```{r}
coffee$NameLength <- str_length(str_trim(coffee$Name))
coffee |> 
  select(Name, NameLength) |> 
  slice(1:10)
```

## Activity: Solutions

3\. Create a new variable in your data for the inspection year.

```{r}
coffee$Year <- str_sub(coffee$`Inspection Date`,-4,-1)
coffee$Year |> 
  head(5)
```

## Activity: Solutions

4\.  Create side-by-side boxplots for the length of business name vs. year.

```{r}
#| fig-height: 6
#| fig-width: 12
#| fig-align: center
library(ggplot2)
ggplot(coffee, aes(Year, NameLength)) + 
  geom_boxplot()
```

## Activity: Solutions

5\. Calculate the maximum `Inspection Score` by business and year.

```{r}
coffee_summary <- coffee %>% 
  group_by(Name, Year) %>% 
  summarize(MaxScore = max(`Inspection Score`))
coffee_summary |> 
  head(5)
```

## Activity: Solutions

6\. Create a line plot of maximum score ("MaxScore") over time ("Year"), by business ("Name"). That is, you should have a single line for each business. (Don't try to label them, as there are far too many!)
```{r}
#| fig-height: 6
#| fig-width: 12
#| fig-align: center
ggplot(coffee_summary, aes(Year, MaxScore, group = Name)) +
  geom_line(alpha = .2)
```

# Homework{.section-title background-color="#99a486"}

## Homework 7

## Due dates

```{r}
#| echo: false
#| message: false
#| warning: false

# reading in base due date schedule
source("../due_dates_schedule.R")

make_due_date_table(7)
```