---
highlight-style:  a11y-dark
footnotes-hover: true
reference-location: margin
execute: 
  echo: true
  message: false
  warning: false
  fig-format: "svg"
format: 
  revealjs:
    theme: lecture_styles.scss
    slide-number: true
    code-link: true
    chalkboard: true
    incremental: false 
    smaller: true
    preview-links: true
    code-line-numbers: true
    history: false
    progress: true
    #logo: "../../images/csss-white.png" # need to add css for size <https://stackoverflow.com/questions/74083521/how-to-increase-size-logo-in-quarto-reveal-js-presentation>
---

```{r}
#| echo: false
#| cache: false
require(downlit)
require(xml2)
require(tidyverse)
```

## {#title-slide data-menu-title="Importing, Exporting, and Cleaning Data" background="#1e4655" background-image="../../images/csss-logo.png" background-position="center top 5%" background-size="50%"}


[Importing, Exporting, and Cleaning Data]{.custom-title}

[CS&SS 508 • Lecture 5]{.custom-subtitle}

[31 October 2023]{.custom-subtitle2}

[Victoria Sass]{.custom-subtitle3}

# Roadmap{.section-title background-color="#99a486"}

---

<br>

Last time, we learned about,

* Types of Data
* Vectors
* Matrices
* Lists

. . . 

Today, we will cover,

* Importing and exporting data
* Reshaping data
* Dates and times

# Importing and Exporting Data{.section-title background-color="#99a486"}

* Data packages
* Imporing data with code
* Importing data by "point-and-click"

## Data Packages

R has a *big* user base.  If you are working with a popular data source, it will often have a devoted R package on *CRAN* or *Github*. 

. . . 

Examples:

* `WDI`: World Development Indicators (World Bank)
* `WHO`: World Health Organization API
* `tidycensus`: Census and American Community Survey
* `quantmod`: financial data from Yahoo, FRED, Google

. . . 

If you have an actual data file, you'll have to import it yourself...

## Delimited Text Files

Besides a package, it's easiest when data is stored in a text file. 

. . . 

An example of a comma-separated values (**.csv**) file is below:

```
"Subject","Depression","Sex","Week","HamD","Imipramine"
101,"Non-endogenous","Second",0,26,NA
101,"Non-endogenous","Second",1,22,NA
101,"Non-endogenous","Second",2,18,4.04305
101,"Non-endogenous","Second",3,7,3.93183
101,"Non-endogenous","Second",4,4,4.33073
101,"Non-endogenous","Second",5,3,4.36945
103,"Non-endogenous","First",0,33,NA
103,"Non-endogenous","First",1,24,NA
103,"Non-endogenous","First",2,15,2.77259
```

## `readr`

R has some built-in functions for importing data, such as `read.table()` and `read.csv()`. 

. . . 

The `readr` package provides similar functions, like `read_csv()`, that have slightly better features:

* Faster!
* Better defaults (e.g. doesn't convert characters to factors)
* A *little* smarter about dates and times
* Loading bars for large files

```{r}
library(readr)
```

## `readr` Importing Example

Let's import some data about song ranks on the Billboard Hot 100 in 2000:

```{r}
billboard_2000_raw <- read_csv(file = "https://raw.githubusercontent.com/vsass/CSSS508/main/Lectures/Lecture5/data/billboard.csv")
```

## Did It Load?

```{r}
library(dplyr)
dim(billboard_2000_raw)
names(billboard_2000_raw) %>% 
  head(20)
```

## Alternate Solution

Import the data manually!

In the upper right-hand pane of RStudio (make sure you're in the Environment tab), select:

`Import Dataset > From Text (readr)`

. . . 

**Once you've imported the data, you can `copy/paste` the import code from the console into your file!!**

This makes the process *reproducible!*

## Importing Other Data Types

* For Excel files (`.xls` or `.xlsx`), use package `readxl`
* For Google Docs Spreadsheets, use package `googlesheets4`
* For Stata, SPSS, and SAS files, use package `haven` (`tidyverse`)
* For Stata, SPSS, and Minitab, use package `foreign`

::: aside
Note: You **won't** keep text formatting, color, comments, or merged cells!!
:::

## Writing Delimited Files

Getting data out of R into a delimited file is very similar to getting it into R:

```{r}
write_csv(billboard_2000_raw, path = "billboard_data.csv")
```

This saved the data we pulled off the web in a file called `billboard_data.csv` in my working directory.

# Reshaping Data {background-image="images/tidyr.png" background-size="contain" background-opacity="0.25" background-position="center" .section-title background-color="#99a486"}

## Initial Spot Checks


First things to check after loading new data:

::: {.incremental}
* Did all the rows/columns from the original file make it in?
    + Check using `dim()` or `str()`
* Are the column names in good shape?
    + Use `names()` to check; fix with `rename()`
* Are there "decorative" blank rows or columns to remove?
    + `filter()` or `select()` out those rows/columns
:::

## Tidy Data

**Tidy data** (aka "long data") are such that:

::: {.incremental}
1. The values for a single observation are in their own row.
2. The values for a single variable are in their own column.
3. There is only one value per cell.
:::

. . . 

Why do we want tidy data?

* **Easier to understand** many rows than many columns
* Required for **plotting** in `ggplot2`
* Required for many types of **statistical procedures** (e.g. hierarchical or mixed effects models)
* Fewer issues with **missing values and "imbalanced"** repeated measures data

## Slightly "Messy" Data

::::{.columns}
:::{.column width="60%"}

| **Program**     | **First Year** | **Second Year** |
|-----------------|-----------:|---------:|
| Evans School    |     10     |    6    |
| Arts & Sciences |      5     |    6    |
| Public Health   |      2     |    3    |
| Other           |      5     |    1    |

:::
:::{.column width="40%"}

* What is an **observation**?
    + A group of students from a program of a given year
    

* What are the **variables**?
    + Program, Year


* What are the **values**?
    + Program: Evans School, Arts & Sciences, Public Health, Other
    + Year: First, Second -- **in column headings. Bad!**
    + Count: **spread over two columns!**
:::
::::

## Tidy Version

::::{.columns}
:::{.column width="50%"}

| **Program**     | **Year** | **Count** |
|-----------------|-----------:|---------:|
| Evans School    |     First |    10   |
| Evans School    |     Second   |    6    |
| Arts & Sciences |     First |    5    |
| Arts & Sciences |     Second   |    6    |
| Public Health   |     First |    2    |
| Public Health   |     Second   |    3    |
| Other           |     First |    5    |
| Other           |     Second   |    1    |
:::
:::{.column width="50%"}
* Each variable is a column.

* Each observation is a row.

* Each cell has a single value.
:::
::::

## Billboard is Just Ugly-Messy

```{r}
#| echo: false
billboard_2000_raw %>% 
  head(10)
```

::: aside
Week columns continue up to `wk76`!
:::

## Billboard

::: {.incremental}
* What are the **observations** in the data?
    + Song on the Billboard chart each week
* What are the **variables** in the data?
    + Year, artist, track, song length, date entered Hot 100, week since first entered Hot 100 (**spread over many columns**), rank during week (**spread over many columns**)
* What are the **values** in the data?
    + e.g. 2000; 3 Doors Down; Kryptonite; 3 minutes 53 seconds; April 8, 2000; Week 3 (**stuck in column headings**); rank 68 (**spread over many columns**)
:::

## `tidyr`

The `tidyr` package provides functions to tidy up data. 

. . . 

Key functions:

* **`pivot_longer()`**: takes a set of columns and pivots them down ("longer") to make two new columns (which you can name yourself): 
    * A `name` column that stores the original column names
    * A `value` with the values in those original columns

. . . 

* **`pivot_wider()`**: inverts `pivot_longer()` by taking two columns and pivoting them up and across ("wider") into multiple columns

. . . 

We're going to focus only on `pivot_longer` here, but know that it can be reversed!

## `pivot_longer()`

This function usually takes three arguments:

1. **cols**: The columns we want to modify
1. **names_to**: New variable name to store original columns
1. **values_to**: New variable name to store original values

## Example of `pivot_longer()`

```{r}
#| code-line-numbers: "|3-5"
library(tidyr)
billboard_2000 <- billboard_2000_raw %>%
  pivot_longer(cols=wk1:wk76,
               names_to ="week",
               values_to = "rank")
billboard_2000 %>% 
  head(5)
```

Now we have a single week column!

## Lots of missing values?!

```{r}
summary(billboard_2000$rank)
```

We don't want to keep the `r sum(is.na(billboard_2000$rank))` rows with missing ranks.

## Pivoting Better: `values_drop_na`

Adding the argument `values_drop_na = TRUE` to `pivot_longer()` will remove rows with missing ranks.

```{r}
#| code-line-numbers: "|5"
billboard_2000 <- billboard_2000_raw %>%
  pivot_longer(cols = wk1:wk76, 
               names_to = "week", 
               values_to = "rank", 
               values_drop_na = TRUE)
summary(billboard_2000$rank)
```

. . . 

No more `NA` values!

```{r}
dim(billboard_2000)
```

And way fewer rows!

## `parse_number()`

The week column is of they type character, but it should be numeric.

```{r}
head(billboard_2000$week)
```

. . . 

`parse_number()` grabs just the numeric information from a character string:

```{r}
#| code-line-numbers: "|2"
billboard_2000 <- billboard_2000 %>%
    mutate(week = parse_number(week))
summary(billboard_2000$week)
```

More sophisticated tools for character strings will be covered later in this course!

# Dates and Times {background-image="images/lubridate.png" background-size="contain" background-opacity="0.25" background-position="center" .section-title background-color="#99a486"}

## `lubridate`

The package `lubridate` (part of the tidyverse!) has a *very large* number of functions you can use!

. . . 

* Converting dates/times between formats (DD-MM-YY to YY-MM-DD)
* Extracting dates/times (day of week, month, leap years, etc.)
* Math with dates/times (time zone conversions, etc.)

. . . 

There's too much to cover all of it, but I'll run through a few examples.

## Dates in `billboard_2000` 

```{r}
billboard_2000 %>% 
  select(date.entered) %>% 
  head(10)
```

## Extracting Year, Month, or Day

```{r }
library(lubridate)
head(billboard_2000$date.entered, 5)
year(billboard_2000$date.entered) %>% 
  head(5)
month(billboard_2000$date.entered) %>% 
  head(5)
day(billboard_2000$date.entered) %>% 
  head(5)
```

## Extracting Weekday, Quarter, and Leap Year Boolean

```{r}
wday(billboard_2000$date.entered) %>% 
  head(5)
quarter(billboard_2000$date.entered) %>% 
  head(5)
leap_year(billboard_2000$date.entered) %>% 
   head(5)
```

## Summary

* Importing/Exporting Data: `readr`
* Reshaping data: `tidyr`
* Dates and times `lubridate`

*Let's take a 10 minute break, then reconvene for an activity!*

## Activity!

In groups of 2-3, you will use the [Billboard data](https://raw.githubusercontent.com/vsass/CSSS508/main/Lectures/Lecture5/data/billboard.csv) to investigate a question:

1. Write down a question of interest that could be studied with this data
     * *Which/how many artists had #1 hits?*
     * *How does rank for each song change over time?*
     * *Is there a relationship between highest rank and length of song?*
2. Make the Billboard data *tidy*, perhaps using the code from this lecture.

3. Perform additional steps (if necessary) to help answer your question:
    * Perhaps using `filter`, `select`, `group_by`, `mutate`, `summarize`, etc.

4. Make a plot or table that answers your question and write down your answer in a sentence.

5. Send me your question, plot/table, and written answer on Canvas

## Example: Question

**Question:** Do songs that hit #1 have a different trajectory than those that don't?

```{r}
billboard_2000_question <- billboard_2000 %>%
    group_by(artist, track) %>%
    mutate(`Weeks at #1` = sum(rank == 1),
           `Peak Rank`   = ifelse(any(rank == 1),
                                  "Hit #1",
                                  "Didn't hit #1")) 
```

::: aside
Note: `any(rank==1)` checks to see if *any* value of `rank` is equal to one for the given `artist` and `track`
:::

## Example Visualization

::: {.panel-tabset}

### Code

```{r}
library(ggplot2)
library(ggthemes)
billboard_trajectories <- 
  ggplot(data = billboard_2000_question,
         aes(x = week, y = rank, group = track,
             color = `Peak Rank`)) +
  geom_line(aes(size = `Peak Rank`), alpha = 0.4) +
  theme_tufte() +
  xlab("Week") + ylab("Rank") +
  scale_color_manual(values = c("black", "red")) +
  scale_size_manual(values = c(0.25, 1)) +
  theme(legend.position = c(0.90, 0.75),
        legend.background = element_rect(fill = "transparent"))
```

### Charts of 2000: Beauty!

```{r}
#| fig-height: 6
#| fig-width: 12
billboard_trajectories
```
:::

::: aside
Songs that reach #1 on the Billboard charts appear to last >20 weeks on the charts, while other songs very rarely make it past that point.
:::

# Homework

## Homework 5

*On [Course Website!*](https://vsass.github.io/CSSS508/Homework/HW5/homework5.html)

## Due Dates

* **Section AA:** 
  + Homework #4 due Tuesday, November 7th (before lecture)
  + Peer grading due Sunday, November 12th (by 4:30pm)

<br>

* **Section AB:** 
  + Homework #4 due Thursday, November 9th (before lab)
  + Peer grading due Tuesday, November 14th (before lecture)
  
  