---
title: "Homework 7"
highlight-style: a11y-dark
---

### [Key](homework7_key.qmd)

Click link above for answers to homework 7. 

### Instructions
Answer each of the following questions. Be sure to display **all your code in the rendered** version (use `echo: true` throughout[^longnote]). 

[^longnote]: You can make this a global option for your whole document by putting it directly in the YAML of your qmd:

    ```{YAML}
    ---
    title: "My Document"
    execute:
      echo: true
    ---
    ```

Use the following code to read in the King County restaurants data: 

```{r}
#| eval: false
load(url("https://github.com/vsass/CSSS508/raw/main/Lectures/Lecture7/data/restaurants.Rdata"))
```

### Exercises

#### Question 1: 

What is the shortest restaurant name? What is the longest restaurant name? 

#### Question 2: 

Read the documentation for [`separate_longer_delim()`](https://tidyr.tidyverse.org/reference/separate_longer_delim.html). Use this function to figure out what 10 "words"^[Anything that is separated by spaces counts as a word for this problem.] are most frequently used for the restaurant names in this data set. 

#### Question 3: 

Which zip code has the highest (meaning **worst**) mean `Grade`?^[Note: To accurately reflect the true average you'll want to weight this by restaurant: i.e. get the average grade by restaurant first, then get the average of those averages by zip code.] For that worst-rated zip code, how has its average rating changed by year?^[There are a few ways to do this. Use the method that makes most sense to you!] 

#### Question 4 (OPTIONAL):

Separate the variable `Phone` into three variables that only contain numerical digits: `area_code`, `first_3`, and `last_4`. 

### Due Dates

```{r}
#| echo: false
#| message: false
#| warning: false

# reading in base due date schedule
source("../../Lectures/due_dates_schedule.R")

make_due_date_table(7) |> 
  tab_options(table.background.color = "#f6f7f9")
```