---
title: "Homework 9"
code-annotations: hover
execute: 
  message: false
  warning: false
output: 
  html_document:
    highlight-style: a11y-dark
---

```{r}
#| echo: false
knitr::opts_chunk$set(comment = ">")

library(tidyverse)
```

### Instructions
Answer each of the following questions. Be sure to display **all your code in the rendered** version (use `echo: true` throughout[^longnote]). 

[^longnote]: You can make this a global option for your whole document by putting it directly in the YAML of your qmd:

    ```{YAML}
    ---
    title: "My Document"
    execute:
      echo: true
    ---
    ```

### Exercises

#### Question 1: 

Compute the number of unique values in each column of `palmerpenguins::penguins`^[You'll need to download the `palmerpenguins` package in order to use `penguins` dataset.].

#### Question 2: 

Compute the mean of every column in `mtcars.`

#### Question 3:

Group `diamonds` by `cut`, `clarity`, and `color` then count the number of observations and compute the mean of each numeric column.

#### Question 4: 

What happens if you use a list of functions in `across()`, but don’t name them? How is the output named?

#### Question 5: 

5. Explain what each step of the following pipeline does. If you haven't seen the function before, look up its help page to learn the specifics of what it does.

```{r}
diamonds |> 
  split(diamonds$cut) |>  # <1> 
  map(\(df) lm(price ~ carat, data = df)) |> 
  map(summary) |> 
  map_dbl("r.squared")
  
```
1. A function from base `R` that does not use tidy evaluation and therefore requires base indexing with `$`

### Due Dates

```{r}
#| echo: false
#| message: false
#| warning: false

# reading in base due date schedule
source("../due_dates_schedule.R")

make_due_date_table_spring(9) |> 
  tab_options(table.background.color = "#f6f7f9")
```